3D Regression Heat Map Analysis of Population Study Data

### 2015-06-23 draft.tex
added Bernhards revision

\paragraph{3D prism as data mini-map.}
\design{
%%
In early prototype versions, the 3D prism acted as starting point for the data analysis without the implementation of a separate 2D view.
%%
Slices were shown using cutaway planes.
%%
This approach was not popular with our domain experts, because the complexity of the visualization overwhelmed them. %did not like the combination of 3D interaction and slicing with representatives.
%%
}
%%
The \emph{3D Regression Heat Map} representation \design{was redesigned to} act as an overview over the whole data set.
%%
Its purpose is not to derive detailed information about data points.
%%
It serves as a function similar to a mini-map, guiding the attention to points of interest in the data.
%%
It also gives context information about adjacent data values when using the 2D heat map.
%%
\design{The distinction between overview and details and demand using two different representations was well received with our domain experts.}
%%
The displayed prism shows values above the matrix diagonal.
%%
For formulas with a dynamic target feature (e.g. exploratory analysis using $Z \sim X + Y$), the color encodes the absolute regression metric values (Fig.~\ref{fig:System}b).
%%
Applying this strategy to a formula containing a static target (e.g. $Cancer \sim X + Y + Z$) yields many occlusions, since the CFS algorithm creates the same feature space for every slice.
%%
For such formulas, the 3D view encodes every data element as absolute difference between its regression metric values and the global mean along the z-axis.
%%
This highlights slices with unusually low or high results (Fig.~\ref{fig:Application}).
%%
\magenta{Variables are ordered the same way in the 2D and 3D heat map to preserve the mental model and make them visually analogous.}

\paragraph{Tackling the disadvantages of 3D information visualization.}
%%
3D information visualizations are often criticized for introducing occlusions and interaction problems.
%%
These are often not balanced out by the advantages of using the third dimension for visual mapping.
%%
We aim to minimize these problems.
%%
Since the regression metric (e.g. $R^2$) values are mapped on data point opacity, large values are highlighted in the prism, guiding the focus to the respective slices.
%%
It also creates a sparse representation, since the majority of regression models yield (depending on the data set and the chosen formula) low $R^2$ values.
%%
Also, the preceding correlation-based feature selection reduces the information space significantly, leading to sparse heat maps.
%%
Overlapping is still an issue, but this way greatly reduced in its effect to the readability of the visualization.
%%

%% INTERACTION
Rotating the 3D heat map is restricted to the y-axis, preserving the mental map to position individual features.
%%
The 3D heat map is always oriented according to the 2D representation, allowing for an easy mental combination of the two representations.
%%
Allowing more degrees of freedom was confusing to our users and also did not add value to the visualization.
%%
Also we provide a zoom functionality using the mouse wheel input.
%%
\paragraph{3D heat map slice selection [\textbf{3.ZF}].}
%%
In order to \emph{Zoom, Filter and Analyze Further}, the user has to navigate to different slices of interest.
%%
We propose two ways to achieve this.

%\begin{itemize}
	%\item
	\textbf{We apply the slicing metaphor from 3D volume data.}
	%%
	In medical volume data renderings, slicing views are very common to view details on a selected plane in the scene.
	%%
	We employ this technique for selecting 3D heat map slices (e.g., by moving a plane via vertical mouse input while pressing the right mouse button).
	%%
	We, however, still display the whole 3D object instead of cutting away information down to the slice position.
	%%
	\design{Early prototypes only provided this method to select a slice of interest, which was inefficient when the user was looking for a specific slice.
	%%
	Hence, an additional method was implemented.
	%%
	}
	%%
	%\item

	\textbf{Selecting the slice using a dropdown menu} provides fast access to plane selections when the user already knows the slices of interest.
	%%
%\end{itemize}
%%
The currently selected slice is denoted with a semi-transparent gray plane.
%%
\design{
Early designs rendered the whole \emph{3D Regression Heat Map}, which made it hard to assess the position of the plane.
%%
%Since the regression metrics are mirrored along the diagonal, we decided to use this space to support better identification of the current slice.
Since the regression metrics are mirrored along the diagonal, the space available from visualizing only the prism generated from the upper half of the heat map diagonal is used to display information about the currently selected plane.
}
%%
The \magenta{regression metric values} are projected on this plane to provide an overlapping-free view.
%%
This allows for a easier to identify the current slice.
%%
\paragraph{2D heat map slice visualization [\textbf{4.DD}].}
%%
The 2D heat map (Fig.~\ref{fig:System}c) shows all values below the matrix diagonal of the current slice.
%%
It creates a optical equivalence with the 3D heat map.
%%
To reduce visual clutter, the 2D view only shows dimensions, which are retrieved through the correlation-based feature selection.
%%
The free space above the matrix diagonal is used to display the 3D heat map.
%%

@@

\section{Implementation} \label{implementation}
\begin{figure}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/implementation}
 \caption{
 The front-end (left) is realized with \emph{HTML5}/\emph{CSS3}/\emph{Javascript} and different Javascript libraries, such as \emph{Angular.js}, \emph{Three.js} and D3.js.
 %%
 The web-server back-end (right) is written using \emph{Node.js}, hosting is carried out using \emph{Heroku}.
 %%
 \emph{R} and \emph{OpenCPU} constitute the statistical back-end (top) to compute the \emph{3D Regression Heat Maps}.
 %%
 Additional statistical back-ends can be attached to the system to decrease the computation time.
 %%
 }
  \label{fig:Implementation}
\end{figure}
%%
We rely on web-based technologies for our prototype.
%%
The ongoing transition of open-science software into the web spawned numerous projects, making state-of-the-art algorithms available in this domain.

%\paragraph{Front-End.}
%%
The front-end is created using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%%
\texttt{Angular.js}\footnote{Open Source; Maintained by Google, \href{https://www.angularjs.org/}{\texttt{angularjs.org}}} abstracts web application into models and views, allowing for a responsive way to combine \texttt{HTML} and \texttt{Javascript}.
%%
It forces developers to write modularized code, which makes the components easier expandable while keeping the code maintainable by including unit tests.
%%
The page layout is handled using \texttt{Twitter Bootstrap}\footnote{Open Source; Maintained by Twitter, \href{http://getbootstrap.com}{\texttt{getbootstrap.com}}}, which also provides a rich set of user interface elements with proper stylings.
%%
The 2D heat map is implemented using \texttt{Data driven Documents} (\texttt{D3.js}), which is popular in information visualization using vector graphics \cite{D3}.
%%
It provides fast and easy methods for binding data to graphical elements.
%%
The 3D plot is created using the WebGL-based \texttt{Threejs}\footnote{Open Source; Originally developed by R. Cabello, \href{http://threejs.org}{\texttt{threejs.org}}} library.
%%
% We experimented with different ways for achieving the cube representation, including volume rendering, cube primitives for each data point and shader-based solutions.
% %%
% Open source volume rendering methods are available but do not satisfy our requirements.
% %%
% Creating a cube primitive for each data point resulted in non-interactive frame-rates for data sets larger than 30 features (creating 30$^3$ cube primitives).
% %%
% Therefore, we decided to use a shader-based solution by rendering the cube as a sprite-based particle system, allowing to customize color and opacity of every data point.
% %%
% It also is the fastest solution that we tested.

%\paragraph{Back-End.}
Two server structures serve as back-end.
%%
The first one is the web-server, which is written in \texttt{Javascript} using \texttt{Node.js}\footnote{Open Source; Maintained by Joyent Inc, \href{http://nodejs.org}{\texttt{nodejs.org}}}, running on Googles V8 Javascript runtime environment.
%%
It is hosted on \texttt{Heroku}\footnote{Owned by Salesforce.com, \href{https://www.heroku.com/}{\texttt{heroku.com}}}, a cloud application platform.
%%

The statistical processors yield the second structure.
%%
They rely on the statistical programming language \texttt{R}.\footnote{Open Source; \href{http://r-project.org}{\texttt{r-project.org}}}
%%
It is widely adopted in the statistical analysis community, yielding a rich support of fast state-of-the-art statistics algorithms as well newly published methods.
%%
\texttt{OpenCPU} is an \texttt{R} package and provides an API for accessing it via HTTP calls \cite{Ooms}.
%%
This way, any computer, which runs \texttt{R} can be turned into a statistical processor for our project.
%%
% The back-end functions necessary for all cube calculations are provided via an \texttt{R} package.
% %%
% It uses multi-core optimization to use all machine CPUs to speed up the calculation process.
% %%
% The server workload balances are managed by the front-end code.

%\paragraph{Access and Source.}
%%
A running instance of the \emph{3D Regression Heat Map} prototype can be found under \href{http://regressioncube.herokuapp.com/}{\texttt{regressioncube.herokuapp.com}}.
%%
The source for the prototype is freely available at \texttt{Github}.\footnote{\texttt{R}-based back-end: \href{https://github.com/paulklemm/regression-cube-r-package}{\\\texttt{github.com/paulklemm/regression-cube-r-package}}}$^{,}$\footnote{Front-End and Node.js Webserver: \href{https://github.com/paulklemm/regression-cube-prototype}{\texttt{\\github.com/paulklemm/regression-cube-prototype}}}
%%
Instructions and code on how to create a setup running the \emph{3D Regression Heat Map} statistical back-end through a \texttt{Ubuntu} server using \texttt{OpenCPU} are referenced in the repository.
%%
The front-end can be deployed using \href{https://www.heroku.com/}{\texttt{www.heroku.com}} by cloning the repository into a Heroku app.
%%
\section{Application} \label{application}
%%
\begin{figure*}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/application}
 \caption{
 The analysis of the numerical and dichotomized target feature depicting liver fat values yields similar results (left).
 %%
 In (a), hotspots for somatometrical features with high correlations were found.
 %%
 High correlations were also found for features depicting \emph{hepatic steatosis} (b).
 %%
 A high correlation between \emph{Interleukin-6}, \emph{hepatic steatosis}, \emph{GGT} and \emph{Lipase} (highlighted using arrows) was revealed during the analysis using the 2D heat map.
 %%
 The hypothesis-free analysis of the breast fat data set (right) w.r.t. the \emph{parenchyma tissue percentage} of the breast displays correlations between \emph{age}, \emph{body fat}, \emph{hip circumference} as well as \emph{menstrual period}.
 }
  \label{fig:Application}
\end{figure*}
%%
In this section, we describe how we applied the \emph{3D Regression Heat Map} to two epidemiological data sets.
%%
The hepatic steatosis data set was analyzed using data mining algorithms, yielding risk groups, which we now analyze further.
%%
Also, we try to reproduce the prior results from this analysis as proof-of-concept of our method.
%%
The breast fat data set is the basis for an explorative analysis w.r.t. the influencing parameters of the parenchyma tissue of the female breast.
%%

Both data sets are unusual for epidemiological analysis regarding their feature extent.
%%
Usually, only a few features depicting a hypothesis are compiled into a data set to assess them using statistical tools.
%%
The herein used data sets comprise several hundred features.
%%
Our method focuses on data exploration and knowledge extraction and requires a wide scope of sociodemographic, medical and lifestyle features.
%%
\subsection{Participants, Setup and Procedure}
%%
The ability of a system to discover knowledge is difficult to measure.
%%
Lam et al.~\cite{Lam2012} propose the \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}) technique, which is focused on the characterization of a systems' ability to generate hypotheses and explore the data in order to extract information.
%%
\emph{VDAR} can be carried out using case studies using thinking-aloud techniques to comprehend the reasoning and thought process of the user.
%%
We employ \emph{VDAR} for analyzing our system.
%%
\paragraph{Participants, setup and procedure.}
%%
We conducted a web-based analysis by using an online-meeting software, which features voice chat as well as screen sharing.
%%
Starting an analysis using these techniques takes about 5-10 minutes of setup time.

@@

%%
We conducted the analysis with three participants.
%%
\emph{KH}, a clinician (10 years of experience) with focus on epidemiological research, is the domain expert for the breast fat data set analysis.
%%
She is a radiologist, responsible for the SHIP-MRI acquisition and also responsible of the mammography analysis.
%%
The hepatic steatosis data set is analyzed by \emph{UN}, a data scientist responsible for prior analysis of the data.
%%
The third participant is \emph{TI}, a statistician with focus on epidemiology (8 years of experience) assesses the statistical reliability of the tool and the underlying methods without a focus on a specific data set.
%%
\subsection{The Hepatic Steatosis Data Set}
We employ the data set used by Niemann et al.~\cite{Niemann2014} to identify predictive features w.r.t. the reversible hepatic steatosis disorder.
%%
The dichotomous target feature is derived from the liver fat concentration measured using MRI scans.
%%
Liver fat concentrations of no more than 10\% are mapped to the `negative' class; values greater than 10\% are mapped to the `positive' class to indicate absence or presence of the disease.
%%
The data set contains labels for 578 participants for which the MRI data was available to that point.
%%
The MRI scans for each subject have been introduced in SHIP-2, so the target feature is only present the last study wave.

%%
Apart from the target feature, the data set contains 199 features, comprising sociodemographic features (gender, age), consumption behavior (e.g., alcohol or tobacco), laboratory data (e.g., sera concentrations), and two features depicting the liver ultrasound.
%%
The acquisition wave is denoted using the appendix; 85 features with appendix \emph{s0} denote their affiliation to SHIP-0 (first study moment), 50 features for \emph{s1} and 55 for \emph{s2}, alongside with 10 time-independent SNPs (DNA base pairs).
%%

%%
Niemann et al.~\cite{Niemann2014} show different class distributions of liver fat concentrations of women and men.
%%
For women, an association between age and liver fat was identified.
%%
An appropriate cut-off value of 52 years was set, yielding the most homogeneous class distribution within the resulting subsets.
%%
Based on these observations, we perform our analysis on three populations: \emph{males}, \emph{females (all ages)} and \emph{females older than 52 years}.
%%
\subsection{The Breast Fat Data Set}
The breast fat data set was compiled to find associations between the parenchyma tissue proportion in the female breast compared to other features in the data.
%%
The ratio between parenchyma and cellular connective tissue (breast density) has been shown to be associated with breast cancer.
%%
Studies describe a four to five times increased risk of getting breast cancer for participants with a breast density above 50\% \cite{Mccormack2006}.

The data comprises 1,186 female subjects (368 from \texttt{SHIP-2}, 818 from \texttt{SHIP-TREND-0} cohort).
%%
It contains 231 dimensions, holding information about somatometric features (e.g., body size or weight) consumption behavior (e.g., alcohol or tobacco), personal and medical history (e.g., occupation or prior diseases), women-specific features (e.g., number of born children or contraception type) as well as mammography features (e.g., fat content or parenchyma tissue proportion to volume).
%%
The latter were derived from MRI data for each subject, which was manually segmented by radiologists.
%%

The data of each cohort was presented as individual \texttt{SPSS} files.
%%
All features related to the mammography attributes were stored in an additional file.
%%
We converted the \texttt{SPSS} data sets to CSV and used \texttt{R} to merge the data sets together using their ID.
%%
All features were renamed to be self-explaining, e.g., \emph{chro\_09a} is now denoted as \emph{Disease\_Osteoporosis}.
%%
This avoids the need of defining a separate data dictionary file for translating the feature names.
%%
\subsection{Case 1: Hypothesis-Driven Analysis of the Hepatic Steatosis Data Set}
%%
We refer to each analysis step with regard to its belonging in the VA-Mantra (recall Sec.~\ref{sec:Workflow}).
%%
The analysis goal was reproducing results with our visual analysis framework that are in accordance to the data mining-based results presented by Niemann et al.~\cite{Niemann2014}.
%%
Therefore, \emph{UN} started the [\textbf{1.AF}] step using the dichotomized MRI fat liver concentration and the formula $mrt\_liverfat\_s2 \sim X + Y + Z$ for \emph{male} subjects.
%%
The [\textbf{2.SI}] step using the 3D heat map locates hotspots at the end of the heat map (Fig.~\ref{fig:Application}a).
%%
The Zoom, Filter and Analyze Further Step [\textbf{3.ZF}] was realized by slicing through the 3D heat map using the mouse input to inspect the hotspots.
%%
Analyzing the 2D heat map [\textbf{4.DD}] revealed high correlations for somatometric features, hepatic steatosis indicator features as well as laboratory values, such as \emph{creatinine} (used as renal retention parameter) and \emph{uric acid} (among others used as gout and diabetes risk factors) magnitudes.
%%
Similar results were present for analyzing the \emph{female} groups.
%%
\emph{UN} could reproduce most results, some features exhibit lower correlations though, e.g., \emph{creatinine} magnitudes.
%%
A slight influence of \emph{age} on the target feature could be observed for women ($R^2$ of $0.09$ for females compared to $0.02$ for males).
%% TODO Feedback of Henry!
Relationships not described by Niemann et al.~\cite{Niemann2014} were found, such as enzymes indicating liver dysfunctions, e.g., \emph{aspartate aminotransferase}.
%
%%
Due to the difference between our regression model approach and the decision tree approach presented by Niemann et al.~\cite{Niemann2014}, a complete matching set of correlating features is not expected.
%%

%%
\paragraph{Analysis of non-discretized target feature.}
%%
Since our method can assess numerical target features, the analysis was conducted again for the non-dichotomized target using the same formula.
%%
The 3D heat map already showed lower $R^2$ values in general.
%%
However, the analysis is now based on linear regression and the $R^2$ values cannot be directly compared.
%%
The correlation hotspots matched with the ones from the dichotomous target, but were generally lower ($R^2$ of $0.37$ for somatometric features as opposed to $0.58$).
%%
We assume that the bias introduced by dichotomizing the fat liver content enforces the findings of liver diseases, while using the numerical features is less expressive.

>>>>

\paragraph{3D prism as data mini-map.}
\design{
%%
In early prototype versions, the 3D prism acted as starting point for the data analysis without the implementation of a separate 2D view.
%%
Slices were shown using cutaway planes.
%%
This approach was not popular among epidemiologists, because the complexity of the visualization overwhelmed them. %did not like the combination of 3D interaction and slicing with representatives.
%%
}
%%
The \emph{3D Regression Heat Map} representation \design{was redesigned to} act as an overview over the whole data set.
%%
%Its purpose is not to derive detailed information about data points.
%%
It serves as a function similar to a mini-map, guiding the attention to points of interest in the data.
%%
It also gives context information about adjacent data values when using the 2D heat map.
%%
\design{The distinction between overview and details on demand using two different representations was well received with our domain experts.}
%%
The displayed prism shows values above the matrix diagonal.
%%
For formulas with a dynamic target feature (e.g. exploratory analysis using $Z \sim X + Y$), the color encodes the absolute regression metric values (Fig.~\ref{fig:System}b).
%%
Applying this strategy to a formula containing a static target (e.g. $Cancer \sim X + Y + Z$) yields many occlusions, since the CFS algorithm creates the same feature space for every slice.
%%
For such formulas, the 3D view encodes every data element as absolute difference between its regression metric values and the global mean along the z-axis.
%%
This highlights slices with unusually low or high results (Fig.~\ref{fig:Application}).
%%
\magenta{Variables are ordered the same way in the 2D and 3D heat map to preserve the mental model and make them visually analogous.}

\paragraph{Tackling the disadvantages of 3D information visualization.}
%%
3D information visualizations are often criticized for introducing occlusions and interaction problems.
%%
These are often not balanced out by the advantages of using the third dimension for visual mapping.
%%
We aim to minimize these problems.
%%
Since the regression metric (e.g. $R^2$) values are mapped on data point opacity, large values are highlighted in the prism, guiding the focus to the respective slices.
%%
The visualization is sparse, since the majority of regression models yield (depending on the data set and the chosen formula) low $R^2$ values.
%%
Also, the preceding correlation-based feature selection reduces the information space significantly, leading to sparse heat maps.
%%
Overlapping is still an issue, but greatly reduced in its effect to the readability of the visualization.
%%

%% INTERACTION
Transformation of the 3D heat map is restricted to the y-axis (horizontal only), preserving the mental map to position individual features.
%%
The 3D heat map is always oriented according to the 2D representation, allowing for an easy mental combination of them.
%%
Allowing more degrees of freedom was confusing to our users and also did not add value to the visualization.
%%
Also we provide a zoom functionality using the mouse wheel input.
%%
\paragraph{3D heat map slice selection [\textbf{3.ZF}].}
%%
In order to \emph{Zoom, Filter and Analyze Further}, the user has to navigate to different slices of interest.
%%
We propose two ways to achieve this.

%\begin{itemize}
	%\item
	\textbf{We apply the slicing metaphor from 3D volume data.}
	%%
	In medical volume renderings, slicing views are common to view details on a selected plane in the scene.
	%%
	We employ this technique for selecting 3D heat map slices (e.g., by moving a plane via vertical mouse input while pressing the right mouse button).
	%%
	We, however, still display the whole 3D object instead of cutting away information. %down to the slice position.
	%%
	\design{Early prototypes only provided this method to select a slice of interest, which was inefficient when the user was looking for a specific slice.
	%%
	Hence, an additional method was implemented.
	%%
	}
	%%
	%\item

	\textbf{Selecting the slice using a dropdown menu} containing the feature names provides fast access to plane selections when the user already knows the slices of interest.
	%%
%\end{itemize}
%%
The currently selected slice is displayed as a semi-transparent gray plane.
%%
\design{
Early prototypes rendered the whole \emph{3D Regression Heat Map}, which made it hard to assess the position of the plane.
%%
%Since the regression metrics are mirrored along the diagonal, we decided to use this space to support better identification of the current slice.
Since the regression metrics are mirrored along the diagonal, the space available from visualizing only the prism generated from the upper half of the heat map diagonal is used to display information about the currently selected plane.
}
%%
The \magenta{regression metric values} are projected on this plane to provide an overlapping-free view.
%%
This allows for a easier to identify the current slice.
%%
\paragraph{2D heat map slice visualization [\textbf{4.DD}].}
%%
The 2D heat map (Fig.~\ref{fig:System}c) shows all values below the matrix diagonal of the current slice.
%%
It creates an optical equivalence with the 3D heat map.
%%
To reduce visual clutter, the 2D view only shows dimensions, which are retrieved through the correlation-based feature selection.
%%
The free space above the matrix diagonal is used to display the 3D heat map.
%%

@@

\section{Implementation} \label{implementation}
\begin{figure}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/implementation}
 \caption{
 The front-end (left) is realized with \emph{HTML5}/\emph{CSS3}/\emph{Javascript} and different Javascript libraries, such as \emph{Angular.js}, \emph{Three.js} and D3.js.
 %%
 The web-server back-end (right) is written using \emph{Node.js}, hosting is carried out using \emph{Heroku}.
 %%
 \emph{R} and \emph{OpenCPU} constitute the statistical back-end (top) to compute the \emph{3D Regression Heat Maps}.
 %%
 Additional statistical back-ends can be attached to the system to decrease the computation time.
 %%
 }
  \label{fig:Implementation}
\end{figure}
%%
We rely on web-based technologies for our prototype.
%%
%The ongoing transition of open-science software into the web spawned numerous projects, making state-of-the-art algorithms available in this domain.
%%
%\paragraph{Front-End.}
%%
The front-end is created using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%%
\texttt{Angular.js}\footnote{Open Source; Maintained by Google, \href{https://www.angularjs.org/}{\texttt{angularjs.org}}} abstracts web application into models and views, allowing for a responsive way to combine \texttt{HTML} and \texttt{Javascript}.
%%
In is easily expandable by forcing developers to write modularized code.
%It forces developers to write modularized code, which makes the components easier expandable. %while keeping the code maintainable by including unit tests.
%%
The page layout is handled using \texttt{Twitter Bootstrap}\footnote{Open Source; Maintained by Twitter, \href{http://getbootstrap.com}{\texttt{getbootstrap.com}}}, which also provides a rich set of user interface elements. %with proper stylings.
%%
The 2D heat map is implemented using \texttt{Data driven Documents} (\texttt{D3.js}), which is popular in information visualization using vector graphics \cite{D3}.
%%
It provides fast and easy methods for binding data to graphical elements.
%%
The 3D plot is created using the WebGL-based \texttt{Threejs}\footnote{Open Source; Originally developed by R. Cabello, \href{http://threejs.org}{\texttt{threejs.org}}} library.
%%
% We experimented with different ways for achieving the cube representation, including volume rendering, cube primitives for each data point and shader-based solutions.
% %%
% Open source volume rendering methods are available but do not satisfy our requirements.
% %%
% Creating a cube primitive for each data point resulted in non-interactive frame-rates for data sets larger than 30 features (creating 30$^3$ cube primitives).
% %%
% Therefore, we decided to use a shader-based solution by rendering the cube as a sprite-based particle system, allowing to customize color and opacity of every data point.
% %%
% It also is the fastest solution that we tested.

%\paragraph{Back-End.}
Two server structures serve as back-end.
%%
The web-server is written in \texttt{Javascript} using \texttt{Node.js}\footnote{Open Source; Maintained by Joyent Inc, \href{http://nodejs.org}{\texttt{nodejs.org}}}, running on Googles V8 Javascript runtime environment.
%%
It is hosted on \texttt{Heroku}\footnote{Owned by Salesforce.com, \href{https://www.heroku.com/}{\texttt{heroku.com}}}, a cloud application platform.
%%

The statistical computations are performed on the second structure.
%%
They rely on the statistical programming language \texttt{R}.\footnote{Open Source; \href{http://r-project.org}{\texttt{r-project.org}}}
%%
It is widely adopted in the statistical analysis community, yielding a rich support of fast state-of-the-art statistics algorithms as well newly published methods.
%%
\texttt{OpenCPU} is an \texttt{R} package and provides an API for accessing it via HTTP calls \cite{Ooms}.
%%
This way, any computer, which runs \texttt{R} can be turned into a statistical processor for our project.
%%
% The back-end functions necessary for all cube calculations are provided via an \texttt{R} package.
% %%
% It uses multi-core optimization to use all machine CPUs to speed up the calculation process.
% %%
% The server workload balances are managed by the front-end code.

%\paragraph{Access and Source.}
%%
A running instance of the \emph{3D Regression Heat Map} prototype can be found under \href{http://regressioncube.herokuapp.com/}{\texttt{regressioncube.herokuapp.com}}.
%%
The source for the prototype is freely available at \texttt{Github}.\footnote{\texttt{R}-based back-end: \href{https://github.com/paulklemm/regression-heatmap-r-package}{\\\texttt{github.com/paulklemm/regression-heatmap-r-package}}}$^{,}$\footnote{Front-End and Node.js Webserver: \href{https://github.com/paulklemm/regression-heatmap-prototype}{\texttt{\\github.com/paulklemm/regression-heatmap-prototype}}}
%%
Instructions and code on how to create a setup running the \emph{3D Regression Heat Map} statistical back-end through a \texttt{Ubuntu} server using \texttt{OpenCPU} are referenced in the repository.
%%
The front-end can be deployed using \href{https://www.heroku.com/}{\texttt{www.heroku.com}} by cloning the repository into a Heroku app.
%%
\section{Application} \label{application}
%%
\begin{figure*}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/application}
 \caption{
 The analysis of the numerical and dichotomized target feature depicting liver fat values yields similar results (left).
 %%
 In (a), hotspots for somatometrical features with high correlations were found.
 %%
 High correlations were also found for features depicting \emph{hepatic steatosis} (b).
 %%
 A high correlation between \emph{Interleukin-6}, \emph{hepatic steatosis}, \emph{GGT} and \emph{Lipase} (highlighted using arrows) was revealed during the analysis using the 2D heat map.
 %%
 The hypothesis-free analysis of the breast fat data set (right) w.r.t. the \emph{parenchyma tissue percentage} of the breast displays correlations between \emph{age}, \emph{body fat}, \emph{hip circumference} as well as \emph{menstrual period}.
 }
  \label{fig:Application}
\end{figure*}
%%
In this section, we describe the application of the \emph{3D Regression Heat Map} to two epidemiological data sets.
%%
The hepatic steatosis data set was analyzed using data mining algorithms, yielding risk groups, which we now analyze further.
%%
Also, we try to reproduce the prior results from this analysis as proof-of-concept of our method.
%%
The breast fat data set is the basis for an explorative analysis w.r.t. the influencing parameters of the parenchyma tissue of the female breast.
%%

Both data sets are unusual for epidemiological analysis regarding their feature extent.
%%
Usually, only a few features depicting a hypothesis are compiled into a data set to assess them using statistical tools.
%%
The herein used data sets comprise several hundred features.
%%
Our method focuses on data exploration and knowledge extraction and requires a wide scope of sociodemographic, medical and lifestyle features.
%%
\subsection{Participants, Setup and Procedure}
%%
The ability of a system to discover knowledge is difficult to measure.
%%
Lam et al.~\cite{Lam2012} propose the \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}) technique, which is focused on the characterization of a systems' ability to generate hypotheses and explore the data in order to extract information.
%%
\emph{VDAR} can be carried out based on case studies using thinking-aloud techniques to comprehend the reasoning and thought process of the user.
%%
We employ \emph{VDAR} for analyzing our system.
%%
\paragraph{Participants, setup and procedure.}
%%
We conducted a web-based analysis by using an online-meeting software, which features voice chat as well as screen sharing.
%%
Starting an analysis using these techniques takes about 5-10 minutes of setup time.

@@

%%
We conducted the analysis with three participants.
%%
\emph{KH}, a clinician (10 years of experience) with focus on epidemiological research, is the domain expert for the breast fat data set.
%%
She is a radiologist, responsible for the SHIP-MRI acquisition and also responsible of the mammography analysis.
%%
The hepatic steatosis data set is analyzed by \emph{UN}, a data scientist responsible for prior analysis of the data.
%%
The third participant is \emph{TI}, a statistician with focus on epidemiology (8 years of experience) assesses the statistical reliability of the tool and the underlying methods without a focus on a specific data set.
%%
\subsection{The Hepatic Steatosis Data Set}
We employ the data set used by Niemann et al.~\cite{Niemann2014} to identify predictive features w.r.t. the reversible hepatic steatosis disorder.
%%
The dichotomous target feature is derived from the liver fat concentration measured using MRI scans.
%%
Liver fat concentrations of no more than 10\% are mapped to the `negative' class; values greater than 10\% are mapped to the `positive' class to indicate absence or presence of the disease.
%%
The data set contains labels for 578 participants. %for which the MRI data was available to that point.
%%
The MRI scans for each subject are only available in SHIP-2. %, so the target feature is only present the last study wave.

%%
Apart from the target feature, the data set contains 199 features, comprising sociodemographic features (gender, age), consumption behavior (e.g., alcohol and tobacco), laboratory data (e.g., sera concentrations), and two features depicting the liver ultrasound.
%%
The acquisition wave is denoted using the appendix; 85 features with appendix \emph{s0} denote their affiliation to SHIP-0 (first study moment), 50 features for \emph{s1} and 55 for \emph{s2}, alongside with 10 time-independent SNPs (DNA base pairs).
%%

%%
Niemann et al.~\cite{Niemann2014} show different class distributions of liver fat concentrations of women and men.
%%
For women, an association between age and liver fat was identified.
%%
An appropriate cut-off value of 52 years was set, yielding the most homogeneous class distribution within the resulting subsets.
%%
Based on these observations, we perform our analysis on three populations: \emph{males}, \emph{females (all ages)} and \emph{females older than 52 years}.
%%
\subsection{The Breast Fat Data Set}
The breast fat data set was compiled to find associations between the parenchyma tissue proportion in the female breast compared to other features in the data.
%%
Breast density is denoted as the ratio between parenchyma and cellular connective tissue and has been shown to be associated with breast cancer.
%%
Studies describe a four to five times increased risk of getting breast cancer for participants with a breast density above 50\% \cite{Mccormack2006}.

The data comprises 1,186 female subjects. %(368 from \texttt{SHIP-2}, 818 from \texttt{SHIP-TREND-0} cohort).
%%
It contains 231 dimensions, holding information about somatometric features (e.g., body size and weight) consumption behavior, personal and medical history (e.g., occupation and prior diseases), women-specific features (e.g., number of born children and contraception type) as well as mammography features (e.g., fat content and parenchyma tissue proportion to volume).
%%
The latter were derived from MRI data for each subject, which was manually segmented by radiologists.
%%

The data of each cohort was presented as individual \texttt{SPSS} files.
%%
All features related to the mammography attributes were stored in an additional file.
%%
We converted the \texttt{SPSS} data sets to CSV and used \texttt{R} to merge the data sets together using their ID.
%%
All features were renamed to be self-explaining, e.g., \emph{chro\_09a} is now denoted as \emph{Disease\_Osteoporosis}.
%%
This avoids the need of defining a separate data dictionary file for translating the feature names.
%%
\subsection{Case 1: Hypothesis-Driven Analysis of the Hepatic Steatosis Data Set}
%%
We refer to each analysis step with regard to its belonging in the VA-Mantra (recall Sec.~\ref{sec:Workflow}).
%%
The analysis goal was reproducing results with our visual analysis framework that are in accordance to the data mining-based results presented by Niemann et al.~\cite{Niemann2014}.
%%
Therefore, \emph{UN} started the [\textbf{1.AF}] step using the dichotomized MRI fat liver concentration and the formula $mrt\_liverfat\_s2 \sim X + Y + Z$ for \emph{male} subjects.
%%
The [\textbf{2.SI}] step using the 3D heat map locates hotspots at the end of the heat map (Fig.~\ref{fig:Application} left).
%%
The Zoom, Filter and Analyze Further Step [\textbf{3.ZF}] was realized by slicing through the 3D heat map using the mouse input to inspect the hotspots.
%%
Analyzing the 2D heat map [\textbf{4.DD}] revealed high correlations for somatometric features, hepatic steatosis indicator features as well as laboratory values, such as \emph{creatinine} (used as renal retention parameter) and \emph{uric acid} (among others used as gout and diabetes risk factors) magnitudes.
%%
Similar results were present for analyzing the \emph{female} groups.
%%
\emph{UN} could reproduce most results, some features exhibit lower correlations though, e.g., \emph{creatinine} magnitudes.
%%
A slight influence of \emph{age} on the target feature could be observed for women ($R^2$ of $0.09$ for females compared to $0.02$ for males).
%% TODO Feedback of Henry!
Relationships not described by Niemann et al.~\cite{Niemann2014} were found, such as enzymes indicating liver dysfunctions, e.g., \emph{aspartate aminotransferase}.
%
%%
Due to the difference between our regression model approach and the decision tree approach presented by Niemann et al.~\cite{Niemann2014}, a complete matching set of correlating features is not expected.
%%

%%
\paragraph{Analysis of non-discretized target feature.}
%%
Since our method can assess numerical target features, the analysis was conducted again for the non-dichotomized target using the same formula.
%%
The 3D heat map showed lower $R^2$ values in general.
%%
However, the analysis is now based on linear regression and the $R^2$ values cannot be compared directly.
%%
The correlation hotspots matched with the ones from the dichotomous target, but were generally lower ($R^2$ of $0.37$ for somatometric features as opposed to $0.58$).
%%
We assume that the bias introduced by dichotomizing the fat liver content enforces the findings of liver diseases, while using the numerical features is less expressive.
